%!TEX root = ../THESIS.tex

\begin{acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

My first words of gratitude go to my supervisor Bob Coecke.
He was the mentor who guided me through to the perks of academia, the architect of the Oxford quantum group in which I learnt so much as well as the guru who indoctrinated me in the science of string diagrams.
After he quit his academic position in the middle of my thesis, he also became the most relaxed and enthusiastic boss in all of quantum industry, creating a new research group from scratch to build upon the work we had started together.
I must also thank my co-supervisor Dan Marsden, who played a crucial role during my masters and the beginning of my DPhil, taming my enthusiasm with his pragmatism and rigour.

Although the words are my own, the work presented in this thesis was of a collective nature.
Thus it is in order to acknowledge my co-authors, first and foremost my academic twin Giovanni de Felice.
I would never have written this thesis without the friendship we have built through more than five years of working, cooking and partying together.
Our QNLP experiments would not have come to life without Konstantinos Meichanetzidis, his physicist mindset and his patience to let us explain our abstract nonsense.
I would also like to thank my academic cousins Richie Yeung and Alex Koziell-Pipe, I trust them to keep the spirit of the quantum group alive.

I am indebted to my examiners Sam Staton and Michael Moortgat for their careful reading of my manuscript, their feedback helped much to improve this thesis.
I am grateful to my college advisor Samson Abramsky for his discreet yet thought-provoking supervision.
Aleks Kissinger and Mehrnoosh Sadrzadeh have also provided me with valuable feedback during my transfer of status.
I want to thank Pawel Sobocinski for inviting me at TalTech where I could present many of the ideas of this thesis, but also for introducing me to string diagrams in the first place in Cali, Columbia over seven years ago.

I am grateful to Simon Harrisson for his generous support through the Wolfson Harrison UK Research Council Quantum Foundation Scholarship.
My DPhil was also supported by the Oxford-DeepMind Graduate Scholarship.
I also thank Cambridge Quantum Computing for taking me as a part-time research scientist.
This brought a new industrial dimension to my research, but also made it possible for me to live with my family in Paris during troubled pandemic times.

On a more personal side, special thanks go to my friends: Tommaso Salvatori who proofread my introduction to NLP, the Wolfson Secret Santa, la Confinerie du 117 and many others who will recognise themselves.
Very special thanks to my family: Astrid, Romeo, my parents Godeleine and Imad, my siblings Raphaël, Maylis and Xavier, my grandfather Etienne (RIP), my grandmothers Véronique and Habiba.
\end{acknowledgements}
